Supplementary Material: Bayesian multi-proxy reconstruction of latitudinal temperature gradients applied to the Early Eocene

### 2023-03-20 manuscript/main.qmd
Work in progress...

---
title: "Integrating geochemical and geoecological data to estimate latitudinal temperature gradients: an early Eocene example"
author:
  - "---"
  - Kilian Eichenseer^1^ and Lewis A. Jones^2^
  - "---"
  - "^1^Department of Earth Sciences, Durham University, South Road, DH1 3LE, Durham, United Kingdom"
  - "^2^Centro de Investigación Mariña, Grupo de Ecoloxía Animal, Departamento de Ecoloxía e Bioloxía Animal, Universidade de Vigo, 36310 Vigo, Spain."
  - "---"
  - "**Corresponding author:** kilian.eichenseer@durham.ac.uk"
abstract: \newpage Abstract
format:
  docx:
    reference-doc: template.docx
    highlight-style: github
    fig-dpi: 900
  pdf:
    documentclass: scrartcl
    papersize: letter
    geometry:
      - top=25.4mm
      - bottom=25.4mm
      - left=25.4mm
      - right=25.4mm
      - heightrounded
bibliography: references.bib
csl: climate-of-the-past.csl
---

@@

@@

The increasing availability of large compilations and databases of deep-time climate data now offers the opportunity for empirical palaeoclimate reconstructions of unprecedented scope [e.g. @Veizer2015; @Hollis2019; @Scotese2021; @Grossman2022; @Judd2022]. To exploit the full potential these data sets offer, two fundamental challenges need to be overcome: Disparate types of proxy data need to be integrated to inform a single climate reconstruction, and large-scale scale climate patterns need to be inferred from localised, often sparse observations. To address these challenges, we present a novel Bayesian model that combines quantitative proxies and palaeoecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled climate data.

Existing palaeoclimate reconstructions are generally based on one of three approaches [see @Burgener2023]: 1) Geochemical proxies such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) provide quantitative palaeotemperature estimates using transfer functions based on experimental work and field observations [e.g. @Liu2009; @Wuchter2004]. 2) Palaeobiological, ecological and lithological proxy data provide climate constraints, but no experimentally determined transfer functions exist to directly infer quantitative climate parameters from these records [e.g. @Wolfe1995; @Kvavcek2007; @Cao2019]. Instead, palaeoclimate signals are inferred based on the climatic conditions of analogous, modern sediments [@Chandra2021], or based on the premise that the climatic requirements of ancient taxa, plant traits or ecological communities were similar to those of their nearest modern relatives [@Peppe2011; @Royer2012; @Salonen2019]. 3) Earth system models offer the most highly resolved and complete climate reconstructions, but ultimately rely on constraints provided by the geological and fossil record [@Lunt2021].

@@

```{=html}
<!--
- Calculate lat gradient
- Calculate global T
[Fig. 4]
-->
```

Modelled sea surface temperature estimates were generated with eq. ?, calculating the sea surface temperatures at any latitude with the parameter estimates of each iteration from the posterior. The median, $2.5^{th}$ and $97.5^{th}$ percentile of temperatures where then taken from all temperature estimates obtained at the latitude of interest.

@@

Estimates of the present-day and early Eocene climatic optimum latitudinal temperature gradient using the Bayesian hierarchical model. The estimated present-day (green line) and early Eocene climatic optimum (purple line) latitudinal temperature gradient. The respective ribbons (green and purple shading) depict the 95% credible intervals of the estimated gradients. The yellow points depict the palaeolatitudinal distribution of geochemical data (e.g. $\delta$^18^O) and their respective point estimate temperature values. The green squares depict the palaeolatitudinal distribution of geoecological data (i.e. coral reefs and mangroves assemblages) and their mean temperature values sampled from their respective normal distributions.
:::

>>>>

---
title: "Bayesian multi-proxy reconstruction of latitudinal temperature gradients: An early Eocene example"
author:
  - "---"
  - Kilian Eichenseer^1^ and Lewis A. Jones^2^
  - "---"
  - "^1^Department of Earth Sciences, Durham University, South Road, DH1 3LE, Durham, United Kingdom"
  - "^2^Centro de Investigación Mariña, Grupo de Ecoloxía Animal, Departamento de Ecoloxía e Bioloxía Animal, Universidade de Vigo, 36310 Vigo, Spain."
  - "---"
  - "**Corresponding author:** kilian.eichenseer@durham.ac.uk"
abstract: \newpage Abstract
format:
  docx:
    reference-doc: template.docx
    highlight-style: github
    fig-dpi: 900
  pdf:
    documentclass: scrartcl
    papersize: letter
    geometry:
      - top=25.4mm
      - bottom=25.4mm
      - left=25.4mm
      - right=25.4mm
      - heightrounded
bibliography: references.bib
csl: climate-of-the-past.csl
---

@@

Palaeoclimate, latitudinal temperature gradients, temperature proxies, Eocene, spatial bias

@@

Reconstructing Earth's past climate is fundamental to understanding how ecosystems respond to climatic and environmental change. To reconstruct palaeoclimatic conditions, researchers use a range of approaches including geochemical proxies, geological evidence, and Earth System Models [see @Burgener2023]. Geochemical proxies such as oxygen isotopes (𝛅^18^O) are used to provide quantitative palaeotemperature estimates by applying transfer functions derived from experimental work and field observations [e.g. @Liu2009; @Wuchter2004]. Geological evidence such as lithological and fossil data are used to qualitatively infer past climatic conditions based on analogous modern sediments [@Chandra2021], or the premise that the climatic requirements of ancient taxa, biological traits, or ecological communities were similar to those of their nearest modern relatives [@Peppe2011; @Royer2012; @Salonen2019]. Finally, Earth System Models (ESMs)--such as the HadCM3 climate model [@valdes2017]--are used to model the main processes of the Earth system to provide spatially and temporally explicit quantitative estimates of palaeoclimatic conditions. However, ESMs ultimately rely on constraints provided by the geological and fossil record [@Lunt2021].

The recent increasing availability of large-scale compilations of deep-time climate proxy data provide an unprecedented opportunity to reconstruct palaeoclimatic conditions [e.g. @Veizer2015; @Hollis2019; @Scotese2021; @Grossman2022; @Judd2022]. However, two fundamental challenges must still be addressed to exploit the full potential of these data sets. First, disparate types of proxy data need to be integrated in order to generate unified palaeoclimatic reconstructions, making use of all available data. Second, the impact of incomplete preservation and uneven sampling of proxy records needs to be addressed [@Jones2022].

To address these challenges, we present a novel Bayesian model that combines quantitative proxies and palaeoecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled climate data.

Routinely, one of three approaches are used to reconstruct past conditions one of three approaches

constrain palaeoclimatic conditions, but no experimentally determined transfer functions exist to directly infer quantitative climate parameters from these records [e.g. @Wolfe1995; @Kvavcek2007; @Cao2019]. Instead, palaeoclimate signals (3)

@@

```{=html}
<!--
- Calculate lat gradient
- Calculate global T
[Fig. 4]
-->
```
Modelled sea surface temperature estimates were generated with eq. ?, calculating the sea surface temperatures at any latitude with the parameter estimates of each iteration from the posterior. The median, $2.5^{th}$ and $97.5^{th}$ percentile of temperatures where then taken from all temperature estimates obtained at the latitude of interest.

@@

Estimates of the present-day and early Eocene climatic optimum latitudinal temperature gradient using the Bayesian hierarchical model. The estimated present-day (green line) and early Eocene climatic optimum (purple line) latitudinal temperature gradient. The respective ribbons (green and purple shading) depict the 95% credible intervals of the estimated gradients. The yellow points depict the palaeolatitudinal distribution of geochemical data (e.g. \$\\delta\$^18^O) and their respective point estimate temperature values. The green squares depict the palaeolatitudinal distribution of geoecological data (i.e. coral reefs and mangroves assemblages) and their mean temperature values sampled from their respective normal distributions.
:::

### 2023-03-20 manuscript/supplementary.qmd
Work in progress...

---
title: "Supplementary Material: Bayesian multi-proxy reconstruction of latitudinal temperature gradients applied to the Early Eocene"
author:
  - "---"
  - Kilian Eichenseer^1^ and Lewis A. Jones^2^
  - "---"
  - "^1^Department of Earth Sciences, Durham University, South Road, DH1 3LE, Durham, United Kingdom"
  - "^2^Centro de Investigación Mariña, Grupo de Ecoloxía Animal, Departamento de Ecoloxía e Bioloxía Animal, Universidade de Vigo, 36310 Vigo, Spain."
  - "---"
  - "**Corresponding author:** kilian.eichenseer@durham.ac.uk"
format:
  docx:
    reference-doc: template.docx
    highlight-style: github
    fig-dpi: 900
  pdf:
    documentclass: scrartcl
    papersize: letter
    geometry:
      - top=25.4mm
      - bottom=25.4mm
      - left=25.4mm
      - right=25.4mm
      - heightrounded
bibliography: references.bib
csl: geology.csl
---

>>>>

---
title: "Supplementary Material: Bayesian multi-proxy reconstruction of latitudinal temperature gradients applied to the Early Eocene"
author:
  - "---"
  - Kilian Eichenseer^1^ and Lewis A. Jones^2^
  - "---"
  - "^1^Department of Earth Sciences, Durham University, South Road, DH1 3LE, Durham, United Kingdom"
  - "^2^Centro de Investigación Mariña, Grupo de Ecoloxía Animal, Departamento de Ecoloxía e Bioloxía Animal, Universidade de Vigo, 36310 Vigo, Spain."
  - "---"
  - "**Corresponding author:** kilian.eichenseer@durham.ac.uk"
format:
  docx:
    reference-doc: template.docx
    highlight-style: github
    fig-dpi: 900
  pdf:
    documentclass: scrartcl
    papersize: letter
    geometry:
      - top=25.4mm
      - bottom=25.4mm
      - left=25.4mm
      - right=25.4mm
      - heightrounded
bibliography: references.bib
csl: climate-of-the-past.csl
---
